---
title: "Titanic: Machine Learning from Disaster"
author: "Abdaljbbar B. A. Dawod"
date: "10th September, 2020"
output:
  html_document:
    df_print: paged
---
To build a prediction model in __R__, several steps should be performed sequentially. This project is structured in a way that will lead us to extract  a prediction model by utilizing the provided datasets for training and testing purposes. Therefore, this project is arranged as:

1. Import CSV-formats datasets into __R__.
2. Investigate the existing structure of the imported dataset.
3. Create appropriate __descriptive statistics__ for the different variables in step __1__.
4. Transform variables that do not provide enough information or we expect additional information if the existing variable has been transformed. 
5. Build the model according to the response variable and explanatory variables, e.g., Linear regression, non-linear regression, poisson regression, logistic regression, etc.
6. Train the model using the train dataset.
7. Estimate the mean squared error of the model.
8. Use the model in (5) to predict the future values of the response variable when the explanatory variables are provided.

Reading dataset into R environment
```{r}
train = read.csv("D:\\Datasets\\Titanic\\train.csv")
```

Show the structure of the loaded dataset
```{r}
str(train)    #the structure of our data
```

# Data transformation.

Changing the Structure of some variables from integer to __factor-type__. 

```{r}
lab = c("No", "Yes")
train$Survived = factor(train$Survived, levels = c(0, 1), labels =lab)
train$Pclass =  as.factor(train$Pclass)
train$Name = as.character(train$Name)
```

Summarizing the loaded dataset.
```{r}
summary(train)
```

Loading the ggplot package for visualization purposes
```{r, warning=FALSE}
library(ggplot2)
```

Plotting  the number of survived passengers _Vs_ dead passengers
```{r}
ggplot(data = train, aes(x = Survived, fill = Survived)) + geom_bar() + ylab("Total Counts")
```

Plotting the survival status passengers according to their class
```{r}
ggplot(data = train, aes(x = Pclass, fill = Survived)) + geom_bar() + ylab("Total Counts")
```

Plotting the survival status of the passengers  according to the gender
```{r}
ggplot(data = train, aes(x = Sex, fill = Survived)) + geom_bar() + ylab("Total Counts")
```

Plotting the survival status of the passangers according to the age
```{r}
ggplot(data = train, aes(y = Age, x = Survived)) + geom_boxplot(fill="lightblue")
```


# Extracting titles from the name variable.

```{r, warning=F}
library(stringr)
ExtractTitle = function(name)
{
  Name = as.character(name)
  if (length(grep("Miss.", Name)) > 0)
    return("Miss.")
  else if (length(grep("Mrs.", Name)) > 0)
    return("Mrs.")
  else if (length(grep("Master.", Name)) > 0)
    return("Master.")
  else if (length(grep("Mr.", Name)) > 0)
    return("Mr.")
  else
    return("Other")
}
```


```{r}
Title = character()
for (i in 1:nrow(train))
  Title = c(Title, ExtractTitle(train$Name[i]))
table(Title)
```

transforming the variable "title"  from string to factor-type.
```{r}
train$Title = as.factor(Title)
```

printing the top 5 rows of train dataset.
```{r}
head(train,5)
```

Extracting boys from train dataset using the variable title. 

```{r}
boys = train[which(train$Title=="Master."),]
```

printing the number of rows of the extracted dataset __boys__.
```{r}
nrow(boys)
```

Summarizing the age of boys
```{r}
summary(boys$Age)
```
Creating age-group variable for the boys dataset
```{r}
boys$Agegroup = cut(boys$Age, breaks = c(0,4,7,12))
```
Printing a table view of the age-group.
```{r}
table(boys$Agegroup)
```

Visualizing the Age of boys 

```{r}
ggplot(data = boys, aes(x = Agegroup, fill=Survived))+
geom_bar() +
ggtitle("Survival status of boys according to their age groups.")+
xlab("Age group")+
ylab("Total count")
```

Extracting __Miss.__ from train dataset using the variable title. 

```{r}
miss = train[which(train$Title=="Miss."),]
```

Creating Age-group variable for the girls dataset
```{r}
miss$Agegroup = cut(miss$Age, breaks = c(0,15,30,40, 80))
```
Printing a table-view of age-groups with their associated counts.
```{r}
table(miss$Agegroup)
```
Plotting Survival status of misses according to their age groups
```{r}
ggplot(data = miss, aes(x = Agegroup, fill=Survived))+
geom_bar() +
ggtitle("Survival status of misses according to their age groups.")+
xlab("Age group")+
ylab("Total count")
```


Building a __logistic regression__ model to predict the probability of survived passanger relying on the passanger class, gender, age, siblings and parents as predictor variables.

```{r}
model = glm(Survived ~ Pclass + Sex + Pclass:Sex + Age + SibSp  , data = train, family = binomial(link = 'logit') )
```

Testing the accuracy of the built model using "test" dataset.
```{r}
test = read.csv("D:\\Datasets\\Titanic\\test.csv") 
test$Pclass =  as.factor(test$Pclass)
test$Name = as.character(test$Name)
summary(model)
anova(model, "Chisq")
```

Predicting the probability of survival status for the first 10 passangers
```{r}
result = as.data.frame(cbind(PassengerId = test$PassengerId, Survived = predict(model, test, type = 'response') ))
cat("Predicting the probability of survival status", sep = "\n", file="predictions.csv")
write.csv(result, file="predictions.csv")
```

Plotting the density of the prediction
```{r}
train$pred <- predict(model, newdata=train, type="response")
test$pred <- predict(model, newdata=test, type="response")
ggplot(train, aes(x=pred, color=Survived, linetype=Survived)) +
geom_density()
```

##  --------------------------__End__ --------------------------                          